Bandit Multiclass Linear Classification: \\ Efficient Algorithms for the Separable Case

### 2019-05-08 ICML-camera-ready/algorithm-for-strongly-linearly-separable-data.tex
add camera-ready folder
pushed the scaling into the proof of Theorem 8
Fixes

In this section, we consider the case when the examples are strongly linearly
separable. We present an algorithm for this setting
(Algorithm~\ref{algorithm:algorithm-for-strongly-linearly-separable-examples})
and give an upper bound on its number of mistakes, stated as
\autoref{theorem:strongly-separable-examples-mistake-upper-bound} below. The
proof of the theorem can be found in
Appendix~\ref{section:proofs-for-stringly-separable-examples} in the
supplementary material.

The idea behind
Algorithm~\ref{algorithm:algorithm-for-strongly-linearly-separable-examples} is
to use $K$ copies of the \textsc{Binary Perceptron} algorithm, one copy per
class; see e.g. \citet[Section 3.3.1]{Shalev-Shwartz-2012}. Upon seeing each
example $x_t$, copy $i$ predicts whether or not $x_t$ belongs to class $i$.
Multiclass predictions are done by evaluating all $K$ binary predictors and
outputting any class with a positive prediction.  If all binary predictions are
negative, the algorithm chooses a prediction a prediction uniformly at random
from $\cbr{1,\ldots,K}$.

@@

\begin{theorem}[Mistake upper bound]
\label{theorem:strongly-separable-examples-mistake-upper-bound}
Let $(V, \ip{\cdot}{\cdot})$ be an inner product space, $K$ be a positive
integer, $\gamma$ be a positive real number, $R$ be a non-negative real number.
If the examples $(x_1, y_1), \dots, (x_T, y_T) \in V \times \{1,2,\dots,K\}$ are
strongly linearly separable with margin $\gamma$ and $\norm{x_1}, \norm{x_2},
\dots, \norm{x_T} \le R$ then the expected number of mistakes
Algorithm~\ref{algorithm:algorithm-for-strongly-linearly-separable-examples}
makes is at most $(K-1) \lfloor 4(R/\gamma)^2 \rfloor$.
\end{theorem}

The upper bound $(K-1) \lfloor 4(R/\gamma)^2 \rfloor$ on the expected number of
mistakes of
Algorithm~\ref{algorithm:algorithm-for-strongly-linearly-separable-examples} is
optimal up to a constant factor, as long as the number of classes $K$ is at most
$O((R/\gamma)^2)$. This lower bound is stated as
\autoref{theorem:strongly-separable-examples-mistake-lower-bound} below. The
proof of the theorem can be found in
Appendix~\ref{section:proofs-for-stringly-separable-examples} in the
supplementary material. \citet{Daniely-Helbertal-2013} provide a lower bound
under the assumption of weak linear separability, which does not immediately
imply a lower bound under the stronger notion.

@@

\begin{theorem}[Mistake upper bound for kernelized algorithm]
\label{theorem:kernelized-upper-bound}
Let $X$ be a non-empty set, let $(V, \ip{\cdot}{\cdot})$ be an inner product
space. Let $\phi:X \to V$ be a feature map and let $k:X \times X \to \R$,
$k(x,x') = \ip{\phi(x)}{\phi(x')}$ be the associated positive definite
kernel. Let $K$ be a positive integer, $\gamma$ be a positive real number, $R$
be a non-negative real number. If $(x_1, y_1), (x_2, y_2), \dots, (x_T, y_T) \in
X \times \{1,2,\dots,K\}$ are labeled examples such that:
\begin{enumerate}
\item the mapped examples $(\phi(x_1), y_1)$, $\dots$, $(\phi(x_T), y_T)$
are strongly linearly separable with margin $\gamma$,
\item $k(x_1, x_1), k(x_2, x_2), \dots, k(x_T,x_T) \le R^2$,
\end{enumerate}
then the expected number of mistakes Algorithm~\ref{algorithm:kernelized} makes
is at most $(K-1) \lfloor 4(R/\gamma)^2 \rfloor$.
\end{theorem}

>>>>

In this section, we consider the case when the examples are strongly linearly
separable. We present an algorithm for this setting
(Algorithm~\ref{algorithm:algorithm-for-strongly-linearly-separable-examples})
and give an upper bound on its number of mistakes, stated as
\autoref{theorem:strongly-separable-examples-mistake-upper-bound} below. The
proof of the theorem can be found in
Appendix~\ref{section:proofs-for-stringly-separable-examples}.

The idea behind
Algorithm~\ref{algorithm:algorithm-for-strongly-linearly-separable-examples} is
to use $K$ copies of the \textsc{Binary Perceptron} algorithm, one copy per
class; see e.g. \citep[Section 3.3.1]{Shalev-Shwartz-2012}. Upon seeing each
example $x_t$, copy $i$ predicts whether or not $x_t$ belongs to class $i$.
Multiclass predictions are done by evaluating all $K$ binary predictors and
outputting any class with a positive prediction.  If all binary predictions are
negative, the algorithm chooses a prediction uniformly at random
from $\cbr{1,2,\dots,K}$.

@@

\begin{theorem}[Mistake upper bound]
\label{theorem:strongly-separable-examples-mistake-upper-bound}
Let $(V, \ip{\cdot}{\cdot})$ be an inner product space, $K$ be a positive
integer, $\gamma$ be a positive real number, $R$ be a non-negative real number.
If the examples $(x_1, y_1), \dots, (x_T, y_T) \in V \times \{1,2,\dots,K\}$ are
strongly linearly separable with margin $\gamma$ and $\norm{x_1}, \norm{x_2},
\dots, \norm{x_T} \le R$ then the expected number of mistakes that
Algorithm~\ref{algorithm:algorithm-for-strongly-linearly-separable-examples}
makes is at most $(K-1) \lfloor 4(R/\gamma)^2 \rfloor$.
\end{theorem}

The upper bound $(K-1) \lfloor 4(R/\gamma)^2 \rfloor$ on the expected number of
mistakes of
Algorithm~\ref{algorithm:algorithm-for-strongly-linearly-separable-examples} is
optimal up to a constant factor, as long as the number of classes $K$ is at most
$O((R/\gamma)^2)$. This lower bound is stated as
\autoref{theorem:strongly-separable-examples-mistake-lower-bound} below. The
proof of the theorem can be found in
Appendix~\ref{section:proofs-for-stringly-separable-examples}.
\citet{Daniely-Helbertal-2013} provide a lower bound
under the assumption of weak linear separability, which does not immediately
imply a lower bound under the stronger notion.

@@

\begin{theorem}[Mistake upper bound for kernelized algorithm]
\label{theorem:kernelized-upper-bound}
Let $X$ be a non-empty set, let $(V, \ip{\cdot}{\cdot})$ be an inner product
space. Let $\phi:X \to V$ be a feature map and let $k:X \times X \to \R$,
$k(x,x') = \ip{\phi(x)}{\phi(x')}$ be the associated positive definite
kernel. Let $K$ be a positive integer, $\gamma$ be a positive real number, $R$
be a non-negative real number. If $(x_1, y_1), (x_2, y_2), \dots, (x_T, y_T) \in
X \times \{1,2,\dots,K\}$ are labeled examples such that:
\begin{enumerate}
\item the mapped examples $(\phi(x_1), y_1)$, $\dots$, $(\phi(x_T), y_T)$
are strongly linearly separable with margin $\gamma$,
\item $k(x_1, x_1), k(x_2, x_2), \dots, k(x_T,x_T) \le R^2$,
\end{enumerate}
then the expected number of mistakes that Algorithm~\ref{algorithm:kernelized} makes
is at most $(K-1) \lfloor 4(R/\gamma)^2 \rfloor$.
\end{theorem}

### 2019-05-08 ICML-camera-ready/appendix-proofs-of-polynomial-approximation-theorems.tex
Fixes

\begin{proof}
If $k = 0$, the inequality trivially holds. For the rest of the proof we can
assume $k \ge 1$. We write the binomial coefficient as
\begin{align*}
\binom{n}{k}
& = \frac{n(n-1)\cdots(n-k+1)}{k(k-1) \cdots 1} \\
& = \frac{n}{k} \cdot \frac{n-1}{k - 1} \cdots \frac{n-k+1}{1} \; .
\end{align*}
We claim that
$$
\frac{n}{k} \le \frac{n-1}{k - 1} \le \cdots \le \frac{n-k+1}{1}
$$
from which the lemma follows by upper bounding all the fractions by $n-k+1$.
It remains to prove that for any $j=0,1,\dots,k-1$,
$$
\frac{n - j + 1}{k - j + 1} \le \frac{n - j}{k - j} \; .
$$
Multiplying by the (positive) denominators, we get an equivalent inequality
$$
(n - j + 1)(k - j) \le (n - j)(k - j + 1) \; .
$$
Cancelling common terms leads to an equivalent inequality
$$
k - j \le n - j \; ,
$$
which since $n \ge k$ by assumption.
\end{proof}

@@

To construct the polynomial $p$ we use Chebyshev polynomials of the first kind.
Chebyshev polynomials of the fist kind form an infinite sequence of polynomials
$T_0(z), T_1(z), T_2(z), \dots$ of single real variable $z$. They are defined
by the recurrence
\begin{align*}
T_0(z) & = 1  \; , \\
T_1(z) & = z  \; ,\\
T_{n+1}(z) & = 2zT_n(z) - T_{n-1}(z), \quad \text{for $n \ge 1$.}
\end{align*}
Chebyshev polynomials have a lot of interesting properties.
We will need properties listed in
\autoref{proposition:properties-of-chebyshev-polynomials} below.
Interested reader can learn more about Chebyshev polynomials
from the book by \cite{Mason-Handscomb-2002}.

@@

The sixth property is equivalent to
$$
T_n(\cosh(\theta)) \ge 1 + n^2 (\cosh(\theta) - 1) \qquad \text{for all $\theta \ge 0$,}
$$
since $\cosh(\theta) = \frac{e^{\theta} + e^{-\theta}}{2}$ is an even continuous
function that maps $\R$ onto $[1,+\infty)$, is strictly decreasing on
$(-\infty,0]$, and is strictly increasing on $[0,\infty)$. Using the fourth
property the last inequality is equivalent to
$$
\cosh(n \theta) \ge 1 + n^2 (\cosh(\theta) - 1) \qquad \text{for all $\theta \ge 0$.}
$$
For $\theta = 0$, both sides are equal to $1$. Thus, it is sufficient to prove
that the derivative of the left hand side is greater or equal to the derivative
of the right hand side. Recalling that $[\cosh(\theta)]' = \sinh(\theta)$, this
means that we need to show that
$$
\sinh(n \theta) \ge n \sinh(\theta) \qquad \text{for all $\theta \ge 0$.}
$$
Tho prove this inequality we use the summation formula
$$
\sinh(\alpha + \beta) = \sinh(\alpha) \cosh(\beta) + \sinh(\beta) \cosh(\beta) \; .
$$
If $\alpha, \beta$ are non-negative then $\sinh(\alpha), \sinh(\beta)$ are
non-negative and $\cosh(\alpha), \cosh(\beta) \ge 1$. Hence,
$$
\sinh(\alpha + \beta) \ge \sinh(\alpha) + \sinh(\beta) \qquad \text{for any $\alpha, \beta \ge 0$.}
$$
This implies that (using induction on $n$) that $\sinh(n \theta) \ge n
\sinh(\theta)$ for all $\theta \ge 0$.

@@

%We prove the following lemma in this section.
%\autoref{theorem:polynomial-approximation-2} immediately
%follows from this lemma by considering $p' = p \cdot 2^{-s(s+1)rm+1}$
%and algebra.

%\begin{lemma}
%\label{lemma:polynomial-approximation-2}
%Let $v_1, v_2, \dots, v_m \in \R^d$ be vectors such that $\norm{v_1},
%\norm{v_2}, \dots, \norm{v_m} \le 1$. Let $\gamma \in (0,1)$.
%Define
%$$
%r = 2 \left\lceil \frac{1}{4} \log_2(4m + 1) \right\rceil + 1 \quad \text{and} \quad s = \left \lceil \log_2(1/\gamma) \right \rceil \; .
%$$
%Then, there exists a multivariate polynomial $p:\R^d \to \R$ such that
%\begin{enumerate}
%\item $\displaystyle p(x) \ge \frac{1}{4} \cdot 2^{s(s+1)rm}$ for all $\displaystyle x \in \bigcap_{i=1}^m \left\{ x \in \R^d ~:~ \norm{x} \le 1, \ \ip{v_i}{x} \ge \gamma \right\}$,

%\item $\displaystyle p(x) \le - \frac{1}{4} \cdot 2^{s(s+1)rm}$ for all $\displaystyle x \in \bigcup_{i=1}^m \left\{ x \in \R^d ~:~ \norm{x} \le 1, \ \ip{v_i}{x} \le - \gamma \right\}$,

%\item $\deg(p) \le (2s+1) rm$,
%\item $\norm{p} \le (2m-1/2) 2^m \cdot \left(2^{2s} rm (4s+2)^2 \right)^{(s+1/2)rm}$.
%\end{enumerate}
%\end{lemma}

We define several univariate polynomials
\begin{align*}
P_n(z) & = (z - 1) \prod_{i=1}^n (z - 2^i)^2, \quad \text{for $n \ge 0$,} \\
A_{n,k}(z) & = (P_n(z))^k - (P_n(-z))^k,  \quad \text{for $n,k \ge 0$,} \\
B_{n,k}(z) & = - (P_n(z))^k - (P_n(-z))^k,  \quad \text{for $n,k \ge 0$.}
\end{align*}
We define the polynomial $p':\R^d \to \R$ as
\[
p'(x) = \left[ \sum_{i=1}^m A_{s,r}\left( \frac{\ip{v_i}{x}}{\gamma} \right) \prod_{\substack{j ~:~ 1 \le j \le m \\ j \neq i}} B_{s,r} \left( \frac{\ip{v_j}{x}}{\gamma} \right) \right]
- \left(m - \frac{1}{2} \right) \prod_{j=1}^m B_{s,r} \left( \frac{\ip{v_j}{x}}{\gamma} \right) \; .
\]
Finally, we define the polynomial $p := 2^{-s(s+1)rm+1} \cdot p'$. We are going to show
that this polynomial $p$ satisfies the required properties.

For convenience we define univariate rational function
\begin{align*}
S_{n,k}(z) & = \frac{A_{n,k}(z)}{B_{n,k}(z)}, \quad \text{for $n,k \ge 0$,}
\end{align*}
and a multivariate rational function
$$
Q(x) = \left( \sum_{i=1}^m S_{s,r}\left( \frac{\ip{v_i}{x}}{\gamma} \right) \right) - \left(m - \frac{1}{2} \right) \; .
$$
It is easy to verify that
$$
p'(x) = Q(x) \prod_{j=1}^m B_{s,r} \left( \frac{\ip{v_j}{x}}{\gamma} \right) \; .
$$

@@

\begin{lemma}[Properties of $S_{n,r}$ and $B_{n,r}$]
\label{lemma:properties-of-s-n-r}
Let $n,m$ be non-negative integers.
Let $r = 2 \left\lceil \frac{1}{4} \log_2(4m + 1) \right\rceil + 1$. Then,
\begin{enumerate}
\item If $z \in [1,2^n]$ then $S_{n,r}(z) \in [1,1+\frac{1}{2m}]$.
\item If $z \in [-2^n, -1]$ then $S_{n,r}(z) \in [-1-\frac{1}{2m}, -1]$.
\item If $z \in [-1,1]$ then $|S_{n,r}(z)| \le 1$.
\item If $z \in [-2^n,2^n]$ then \\ $B_{n,r}(z) \ge \left(1 - \frac{1}{4m+1} \right) 2^{n(n+1)r}$.
\end{enumerate}
\end{lemma}

@@

%\frac{\displaystyle 1 + \left( - \frac{P_n(z)}{P_n(-z)}\right)^r}{\displaystyle 1 - \left( - \frac{P_n(z)}{P_n(-z)}\right)^r} =

@@

It remains to prove part 4 of \autoref{theorem:polynomial-approximation-2}.
For any $i \in \{0,1,2,\dots,s\}$ and any $v \in \R^d$ such that $\norm{v} \le 1$
define multivariate polynomials
\begin{align*}
f_{i,v}(x) & = \frac{\ip{v}{x}}{\gamma} - 2^i \; , \\
q_v(x) & = P_s \left( \frac{\ip{v}{x}}{\gamma} \right) \; , \\
a_v(x) & = A_{s,r} \left( \frac{\ip{v}{x}}{\gamma} \right) \; , \\
b_v(x) & = B_{s,r} \left( \frac{\ip{v}{x}}{\gamma} \right) \; .
\end{align*}
Note that
\[
p'(x) = \left[ \sum_{i=1}^m a_{v_i}(x) \prod_{\substack{j ~:~ 1 \le j \le m \\ j \neq i}} b_{v_j}(x) \right] - \left(m - \frac{1}{2} \right) \prod_{j=1}^n b_{v_j}(x) \; .
\]
We bound the norms of these polynomials. We have
$$
\norm{f_{i,v}}^2 = \norm{v}^2/\gamma^2 + 2^{2i} \le 2 \cdot 2^{2s} \; .
$$
where we used that $1/\gamma \le 2^s$ and $\norm{v} \le 1$.
Since $q_v(x) = f_{i,v}(\frac{\ip{v}{x}}{\gamma}) \prod_{i=1}^s \left(f_{i,v}(\frac{\ip{v}{x}}{\gamma})\right)^2$,
using part 1 of \autoref{lemma:properties-of-norm-of-polynomials} we upper bound the norm of $q_v$
as
\begin{align*}
\norm{q_v}^2
& \le (2s+1)^{2s+1} \norm{f_{0,v}}^2 \prod_{i=1}^s \norm{f_{i,v}}^4 \\
& \le  (2s+1)^{2s+1} (2 \cdot 2^{2s})^{2s + 1} \; .
\end{align*}
Using parts 3 and 2 of \autoref{lemma:properties-of-norm-of-polynomials} we upper bound the norm of $a_v$ as
\begin{align*}
\norm{a_v}^2
& \le 2\norm{(q_v)^r}^2 + 2\norm{(q_{-v})^r}^2 \\
& \le 2 r^{r(2s+1)} (\norm{q_v}^{2})^r + 2 r^{r(2s+1)} (\norm{q_{-v}}^{2})^r \\
& \le 4 r^{r(2s+1)} \left((2s+1)^{2s+1} (2 \cdot 2^s)^{2s + 1} \right)^{r} \\
& = 4 \left(2^{2s} r (4s+2) \right)^{(2s+1)r} \; .
\end{align*}
The same upper bound holds for $\norm{b_v}^2$. Therefore,
\begin{align*}
\norm{p'}
& \le \left[ \sum_{i=1}^m \norm{a_{v_i} \prod_{\substack{j ~:~ 1 \le j \le m \\ j \neq i}} b_{v_j}} \right] + \left(m - \frac{1}{2} \right) \norm{\prod_{j=1}^m b_{v_j}} \\
& \le \left[ \sum_{i=1}^m m^{(s+1/2)rm} \norm{a_{v_i}} \prod_{\substack{j ~:~ 1 \le j \le m \\ j \neq i}} \norm{b_{v_j}} \right] \\
& \qquad + \left(m - \frac{1}{2} \right) m^{(s+1/2)rm} \prod_{j=1}^m \norm{b_{v_j}} \\
& \le (2m-1/2) m^{(s+1/2)rm} \left(4 \left(2^{2s} r (4s+2) \right)^{(2s+1)r} \right)^{m/2} \\
& = (2m-1/2) 2^m \cdot \left(2^{2s} rm (4s+2) \right)^{(s+1/2)rm} \; .
\end{align*}
Finally,
$\norm{p} = 2^{-s(s+1)rm+1} \norm{p'} \leq (4m-1) 2^m \cdot \left(2^s rm (4s+2) \right)^{(s+1/2)rm}$. The theorem follows.

>>>>

\begin{proof}
If $k = 0$, the inequality trivially holds. For the rest of the proof we can
assume $k \ge 1$. We write the binomial coefficient as
\begin{align*}
\binom{n}{k}
& = \frac{n(n-1)\cdots(n-k+1)}{k(k-1) \cdots 1} \\
& = \frac{n}{k} \cdot \frac{n-1}{k - 1} \cdots \frac{n-k+1}{1} \; .
\end{align*}
We claim that
$$
\frac{n}{k} \le \frac{n-1}{k - 1} \le \cdots \le \frac{n-k+1}{1}
$$
from which the lemma follows by upper bounding all the fractions by $n-k+1$.
It remains to prove that for any $j=0,1,\dots,k-1$,
$$
\frac{n - j + 1}{k - j + 1} \le \frac{n - j}{k - j} \; .
$$
Multiplying by the (positive) denominators, we get an equivalent inequality
$$
(n - j + 1)(k - j) \le (n - j)(k - j + 1) \; .
$$
We multiply out the terms and get
$$
nk - kj + k - nj + j^2 - j \le nk - nj + n - kj + j^2 - j \; .
$$
We cancel common terms and get an equivalent inequality $k \ge n$, which
holds by the assumption.
\end{proof}

@@

To construct the polynomial $p$ we use Chebyshev polynomials of the first kind.
Chebyshev polynomials of the fist kind form an infinite sequence of polynomials
$T_0(z), T_1(z), T_2(z), \dots$ of single real variable $z$. They are defined
by the recurrence
\begin{align*}
T_0(z) & = 1  \; , \\
T_1(z) & = z  \; ,\\
T_{n+1}(z) & = 2zT_n(z) - T_{n-1}(z), \quad \text{for $n \ge 1$.}
\end{align*}
Chebyshev polynomials have a lot of interesting properties.
We will need properties listed in
\autoref{proposition:properties-of-chebyshev-polynomials} below.
Interested reader can learn more about Chebyshev polynomials
from the book by \citet{Mason-Handscomb-2002}.

@@

The sixth property is equivalent to
$$
T_n(\cosh(\theta)) \ge 1 + n^2 (\cosh(\theta) - 1) \qquad \text{for all $\theta \ge 0$,}
$$
since $\cosh(\theta) = \frac{e^{\theta} + e^{-\theta}}{2}$ is an even continuous
function that maps $\R$ onto $[1,+\infty)$, is strictly decreasing on
$(-\infty,0]$, and is strictly increasing on $[0,\infty)$. Using the fourth
property the last inequality is equivalent to
$$
\cosh(n \theta) \ge 1 + n^2 (\cosh(\theta) - 1) \qquad \text{for all $\theta \ge 0$.}
$$
For $\theta = 0$, both sides are equal to $1$. Thus, it is sufficient to prove
that the derivative of the left hand side is greater or equal to the derivative
of the right hand side. Recalling that $[\cosh(\theta)]' = \sinh(\theta)$, this
means that we need to show that
$$
\sinh(n \theta) \ge n \sinh(\theta) \qquad \text{for all $\theta \ge 0$.}
$$
To prove this inequality we use the summation formula
$$
\sinh(\alpha + \beta) = \sinh(\alpha) \cosh(\beta) + \sinh(\beta) \cosh(\beta) \; .
$$
If $\alpha, \beta$ are non-negative then $\sinh(\alpha), \sinh(\beta)$ are
non-negative and $\cosh(\alpha), \cosh(\beta) \ge 1$. Hence,
$$
\sinh(\alpha + \beta) \ge \sinh(\alpha) + \sinh(\beta) \qquad \text{for any $\alpha, \beta \ge 0$.}
$$
This implies that (using induction on $n$) that $\sinh(n \theta) \ge n
\sinh(\theta)$ for all $\theta \ge 0$.

@@

We define several univariate polynomials
\begin{align*}
P_n(z) & = (z - 1) \prod_{i=1}^n (z - 2^i)^2, \quad \text{for $n \ge 0$,} \\
A_{n,k}(z) & = (P_n(z))^k - (P_n(-z))^k,  \quad \text{for $n,k \ge 0$,} \\
B_{n,k}(z) & = - (P_n(z))^k - (P_n(-z))^k,  \quad \text{for $n,k \ge 0$.}
\end{align*}
We define the polynomial $q:\R^d \to \R$ as
$$
q(x) = \left[ \sum_{i=1}^m A_{s,r}\left( \frac{\ip{v_i}{x}}{\gamma} \right) \prod_{\substack{j ~:~ 1 \le j \le m \\ j \neq i}} B_{s,r} \left( \frac{\ip{v_j}{x}}{\gamma} \right) \right]
- \left(m - \frac{1}{2} \right) \prod_{j=1}^m B_{s,r} \left( \frac{\ip{v_j}{x}}{\gamma} \right) \; .
$$
Finally, we define $p(x) = 2^{-s(s+1)rm+1} q(x)$. We are going to show
that this polynomial $p$ satisfies the required properties.

For convenience we define univariate rational function
\begin{align*}
S_{n,k}(z) & = \frac{A_{n,k}(z)}{B_{n,k}(z)}, \quad \text{for $n,k \ge 0$,}
\end{align*}
and a multivariate rational function
$$
Q(x) = \left( \sum_{i=1}^m S_{s,r}\left( \frac{\ip{v_i}{x}}{\gamma} \right) \right) - \left(m - \frac{1}{2} \right) \; .
$$
It is easy to verify that
$$
q(x) = Q(x) \prod_{j=1}^m B_{s,r} \left( \frac{\ip{v_j}{x}}{\gamma} \right) \; .
$$

@@

\begin{lemma}[Properties of $S_{n,r}$ and $B_{n,r}$]
\label{lemma:properties-of-s-n-r}
Let $n,m$ be non-negative integers.
Let $r = 2 \left\lceil \frac{1}{4} \log_2(4m + 1) \right\rceil + 1$. Then,
\begin{enumerate}
\item If $z \in [1,2^n]$ then $S_{n,r}(z) \in [1,1+\frac{1}{2m}]$.
\item If $z \in [-2^n, -1]$ then $S_{n,r}(z) \in [-1-\frac{1}{2m}, -1]$.
\item If $z \in [-1,1]$ then $|S_{n,r}(z)| \le 1$.
\item If $z \in [-2^n,2^n]$ then $B_{n,r}(z) \ge \left(1 - \frac{1}{4m+1} \right) 2^{n(n+1)r}$.
\end{enumerate}
\end{lemma}

@@

@@

It remains to prove part 4 of \autoref{theorem:polynomial-approximation-2}.
For any $i \in \{0,1,2,\dots,s\}$ and any $v \in \R^d$ such that $\norm{v} \le 1$
define multivariate polynomials
\begin{align*}
f_{i,v}(x) & = \frac{\ip{v}{x}}{\gamma} - 2^i \; , \\
q_v(x) & = P_s \left( \frac{\ip{v}{x}}{\gamma} \right) \; , \\
a_v(x) & = A_{s,r} \left( \frac{\ip{v}{x}}{\gamma} \right) \; , \\
b_v(x) & = B_{s,r} \left( \frac{\ip{v}{x}}{\gamma} \right) \; .
\end{align*}
Note that
$$
q(x) = \left[ \sum_{i=1}^m a_{v_i}(x) \prod_{\substack{j ~:~ 1 \le j \le m \\ j \neq i}} b_{v_j}(x) \right] - \left(m - \frac{1}{2} \right) \prod_{j=1}^n b_{v_j}(x) \; .
$$
We bound the norms of these polynomials. We have
$$
\norm{f_{i,v}}^2 = \norm{v}^2/\gamma^2 + 2^{2i} \le 2 \cdot 2^{2s} \; .
$$
where we used that $1/\gamma \le 2^s$ and $\norm{v} \le 1$.
Since $q_v(x) = f_{i,v}(\frac{\ip{v}{x}}{\gamma}) \prod_{i=1}^s \left(f_{i,v}(\frac{\ip{v}{x}}{\gamma})\right)^2$,
using part 1 of \autoref{lemma:properties-of-norm-of-polynomials} we upper bound the norm of $q_v$
as
\begin{align*}
\norm{q_v}^2
& \le (2s+1)^{2s+1} \norm{f_{0,v}}^2 \prod_{i=1}^s \norm{f_{i,v}}^4 \\
& \le  (2s+1)^{2s+1} (2 \cdot 2^{2s})^{2s + 1} \; .
\end{align*}
Using parts 3 and 2 of \autoref{lemma:properties-of-norm-of-polynomials} we upper bound the norm of $a_v$ as
\begin{align*}
\norm{a_v}^2
& \le 2\norm{(q_v)^r}^2 + 2\norm{(q_{-v})^r}^2 \\
& \le 2 r^{r(2s+1)} (\norm{q_v}^{2})^r + 2 r^{r(2s+1)} (\norm{q_{-v}}^{2})^r \\
& \le 4 r^{r(2s+1)} \left((2s+1)^{2s+1} (2 \cdot 2^s)^{2s + 1} \right)^{r} \\
& = 4 \left(2^{2s} r (4s+2) \right)^{(2s+1)r} \; .
\end{align*}
The same upper bound holds for $\norm{b_v}^2$. Therefore,
\begin{align*}
\norm{q}
& \le \left[ \sum_{i=1}^m \norm{a_{v_i} \prod_{\substack{j ~:~ 1 \le j \le m \\ j \neq i}} b_{v_j}} \right] + \left(m - \frac{1}{2} \right) \norm{\prod_{j=1}^m b_{v_j}} \\
& \le \left[ \sum_{i=1}^m m^{(s+1/2)rm} \norm{a_{v_i}} \prod_{\substack{j ~:~ 1 \le j \le m \\ j \neq i}} \norm{b_{v_j}} \right] \\
& \qquad + \left(m - \frac{1}{2} \right) m^{(s+1/2)rm} \prod_{j=1}^m \norm{b_{v_j}} \\
& \le (2m-1/2) m^{(s+1/2)rm} \left(4 \left(2^{2s} r (4s+2) \right)^{(2s+1)r} \right)^{m/2} \\
& = (2m-1/2) 2^m \cdot \left(2^{2s} rm (4s+2) \right)^{(s+1/2)rm} \; .
\end{align*}
Finally,
$\norm{p} = 2^{-s(s+1)rm+1} \norm{q} \leq (4m-1) 2^m \cdot \left(2^s rm (4s+2) \right)^{(s+1/2)rm}$. The theorem follows.

### 2019-05-08 ICML-camera-ready/mistake-lower-bound-for-ignorant-algorithms.tex
Fixes

To formally define ignorant algorithms, we define the conditional distribution
from which an algorithm draws its predictions. Formally, given an algorithm
$\calA$ and an adversarial strategy, we define
\[
p_t(y|x) = \\
\Pr[y_t = y ~|~ (x_1, y_1), (x_2, y_2) \dots, (x_{t-1}, y_{t-1}), x_t = x] \; .
\]
In other words, in any round $t$, conditioned on the past $t-1$ rounds, the
algorithm $\calA$ chooses $y_t$ from probability distribution $p_t(\cdot|x_t)$.
Formally, $p_t$ is a function $p:\{1,2,\dots,K\} \times \R^d \to [0,1]$
such that $\sum_{y=1}^K p_t(y|x) = 1$ for any $x \in \R^d$.

@@

To explain the definition, consider an ignorant algorithm $\calA$. Suppose that
on a sequence of examples $(x_1, y_1)$, $(x_2, y_2)$, $\dots$, $(x_{t-1}, y_{t-1})$
generated by some adversary the algorthm $\calA$ makes correct predictions in
rounds $a_1, a_2, \dots, a_n$ where $1 \le a_1 < a_2 < \dots < a_n < t$ and
errors on rounds $\{1,2,\dots,t-1\} \setminus \{a_1, a_2, \dots, a_n\}$. Suppose
that on another sequence of examples $(x_1', y_1'), (x_2', y_2'), \dots,
(x_{s-1}', y_{s-1}')$ generated by another adversary the algorithm $\calA$ makes
correct predictions in rounds $b_1, b_2, \dots, b_n$ where $1 \le b_1 < b_2 <
\dots < b_n < s$ and errors on rounds $\{1,2,\dots,s-1\} \setminus \{b_1, b_2,
\dots, b_n\}$. Futhermore, suppose
\begin{align*}
(x_{a_1}, y_{a_1}) &= (x'_{b_1}, y'_{b_1}) \; , \\
(x_{a_2}, y_{a_2}) &= (x'_{b_2}, y'_{b_2}) \; , \\
\vdots \\
(x_{a_n}, y_{a_n}) &= (x'_{b_2}, y'_{b_n}) \; .
\end{align*}
Then, as $\calA$ is ignorant,
\begin{align*}
\Pr[y_t = y ~|~ (x_1, y_1), (x_2, y_2) \dots, (x_{t-1}, y_{t-1}), x_t = x] = \\
\Pr[y_t' = y ~|~ (x_1', y_1'), (x_2', y_2') \dots, (x_{t-1}', y_{t-1}'), x_t' = x].
\end{align*}
Note that the sequences $(x_1, y_1)$, $(x_2, y_2)$, $\dots$, $(x_{t-1},
y_{t-1})$ and $(x_1', y_1')$, $(x_2', y_2')$, $\dots$, $(x_{s-1}', y_{s-1}')$
might have different lengths and and $\calA$ might error in different sets of
rounds. As a special case, if an ignorant algorithm makes a mistake in round $t$
then $p_{t+1}=p_t$.

>>>>

To formally define ignorant algorithms, we define the conditional distribution
from which an algorithm draws its predictions. Formally, given an algorithm
$\calA$ and an adversarial strategy, we define
\[
p_t(y|x) =
\Pr[y_t = y ~|~ (x_1, y_1), (x_2, y_2) \dots, (x_{t-1}, y_{t-1}), x_t = x] \; .
\]
In other words, in any round $t$, conditioned on the past $t-1$ rounds, the
algorithm $\calA$ chooses $y_t$ from probability distribution $p_t(\cdot|x_t)$.
Formally, $p_t$ is a function $p:\{1,2,\dots,K\} \times \R^d \to [0,1]$
such that $\sum_{y=1}^K p_t(y|x) = 1$ for any $x \in \R^d$.

@@

To explain the definition, consider an ignorant algorithm $\calA$. Suppose that
on a sequence of examples $(x_1, y_1)$, $(x_2, y_2)$, $\dots$, $(x_{t-1}, y_{t-1})$
generated by some adversary the algorithm $\calA$ makes correct predictions in
rounds $a_1, a_2, \dots, a_n$ where $1 \le a_1 < a_2 < \dots < a_n < t$ and
errors on rounds $\{1,2,\dots,t-1\} \setminus \{a_1, a_2, \dots, a_n\}$. Suppose
that on another sequence of examples $(x_1', y_1'), (x_2', y_2'), \dots,
(x_{s-1}', y_{s-1}')$ generated by another adversary the algorithm $\calA$ makes
correct predictions in rounds $b_1, b_2, \dots, b_n$ where $1 \le b_1 < b_2 <
\dots < b_n < s$ and errors on rounds $\{1,2,\dots,s-1\} \setminus \{b_1, b_2,
\dots, b_n\}$. Futhermore, suppose
\begin{align*}
(x_{a_1}, y_{a_1}) &= (x'_{b_1}, y'_{b_1}) \; , \\
(x_{a_2}, y_{a_2}) &= (x'_{b_2}, y'_{b_2}) \; , \\
\vdots \\
(x_{a_n}, y_{a_n}) &= (x'_{b_2}, y'_{b_n}) \; .
\end{align*}
Then, as $\calA$ is ignorant,
$$
\Pr[y_t = y ~|~ (x_1, y_1), (x_2, y_2) \dots, (x_{t-1}, y_{t-1}), x_t = x] =
\Pr[y_t' = y ~|~ (x_1', y_1'), (x_2', y_2') \dots, (x_{t-1}', y_{t-1}'), x_t' = x].
$$
Note that the sequences $(x_1, y_1)$, $(x_2, y_2)$, $\dots$, $(x_{t-1},
y_{t-1})$ and $(x_1', y_1')$, $(x_2', y_2')$, $\dots$, $(x_{s-1}', y_{s-1}')$
might have different lengths and and $\calA$ might error in different sets of
rounds. As a special case, if an ignorant algorithm makes a mistake in round $t$
then $p_{t+1}=p_t$.
